### 2009-12-12 docs/marcello-dissertation/thesis/chapter6-data-persistence-component-design-and-architecture.tex
Revision containing the English review from Ivan, and rework in most of the sections.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@697 559edde2-a4d4-44e0-96fc-4035e40d7396

In order to provide information to users, the RTC scientists use OPeNDAP
\cite{opendap} server to distribute collected data over the Internet. As
discussed in the previous chapter, this work proposes a data persistence
application of a Key-Value database system in order to provide an alternative
method for data persistence in sensor networks, taking into account the
taxonomies discussed earlier in chapter \ref{chap:taxonomies}.

Considering our case study being NetBEAMS, this chapter describes the design
and architecture component designed specifically for the Data Sensor Platform,
DSP, in order to provide a persistence layer for NetBEAMS. It is vital to note
that the technology chosen in the last chapter can be adapted by any
alternative that uses the data model and infrastructure options presented.

The proposed DSP Data Persistence component for the DSP
Platform was implemented by using the same approaches described in chapter
\ref{chap:netbeams-overview}. The following sections describe the scenarios for
persistence, analyze the requirements from RTC, and describe the design and
architecture of the solution.

@@

As described in section \ref{chap:netbeams-overview}, the DSP Platform was
designed to support an environmental sensor network based in San Francisco Bay,
composed by different types of sensors. The process of extracting the
measurements from sensors can be summarized following the UML Business Process
Diagram \cite{uml} depicted in figure
\ref{fig:dsp-persistence-business-process}. It is important to note that this
diagram details remote communication between sensor devices and a centralized
server, as part of the main scenario to solve the problem of data persistence.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-DataPersistence-Business-Diagram}
  \caption{UML Business Process Model diagram - Adding Persistence for NetBEAMS}
  \label{fig:dsp-persistence-business-process}
\end{figure}

At the event of data collection, the DSP Client receives measurements collected
from each of the sensor devices. Each of the co-located DSP Client process the
collected measurements and transform them into the alleged DSP Message
(section \ref{sec:dsp-message}), which embodies the actual data values. After
proper data processing, transformation and categorization, the DSP Message is
added into temporary outbound queue. Then, as an asynchronous data transmission
event occurs with a specified rate, transferring all the queued messages to be
processed by a DSP Server host. As image
\ref{fig:dsp-persistence-business-process}, the server is centralized and able
to concurrently process messages.

This work proposes the addition of a database system that is responsible for
storing the gathered measurement data from sensor nodes of the SF-BEAMS through
the use of NetBEAMS. Also in figure \ref{fig:dsp-persistence-business-process},
the processed measurement data from sensors 1 through N must be collected in
the same style and categorized as necessary. In this way, end users such as
Biologists or Programmers can access the collected measurements directly
from the Database. Furthermore, the same collected data can be exported to
different formats as necessary.

As a matter of fact, the data measurements collected from the sensor devices
are basic properties with keys and associated values as described in section 
\ref{sec:sfbeams}, after data has been described. According to the descriptions
of the DSP Platform and the DSP Messages structure, the data is represented in
programming language formats such as Java \cite{java} or in a serialized
version in XML \cite{xml}. Similarly, the collected data contains measurements
that uniquely identifies them, together with other properties such as
timestamps of data generation and collected and the IP address of the DSP
Client that produced the data. Therefore, these are essential properties for
the proposed solution of this work.

@@

By analyzing the main scenario of collected data in the previous section, as
well as by collected requirements from the RTC staff, this section lists the
main use cases that the system and the primary users require.

@@

As part of the user stories related to data access, figure
\ref{fig:DSP-Data-Persistence-UseCases-Diagram-Users} describes the use cases
that the end users can perform with the system.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-Data-Persistence-UseCases-Diagram-Users}
  \caption{UML Use Case diagram for Persistence Functions}
  \label{fig:DSP-Data-Persistence-UseCases-Diagram-Users}
\end{figure}

\begin{itemize}
  \item \textbf{Search Measurements by Sensor Properties}: users of the
  collected data must be able to search data through the properties of the
  sensors. This is related to identity principle of what was collected
  as provenance \cite{db-provenance};
  \item \textbf{Search Measurements by Sensor Location}: users of the collected
  data must be able to search data by a given sensor location, as referred to
  the provenance from where the data was collected;
  \item \textbf{Search Measurements by Data collection date}: users of the
  collected data must be able to search data by a given date, or ranges of
  start and end date;
  \item \textbf{Access the data through API, Programming Languages}: users of the
  collected data must be able to search data by the use of different
  programming and scripting languages of their preference;
  \item \textbf{Export Data to OPEnDAP format}: users of the collected data
  must be able to export the collected data to the OPEnDAP format;
  \item \textbf{Export Data to Spreadsheets}: users of the collected data must
  be able to export the collected data to spreadsheets.
\end{itemize}

@@

Different constraints are required to execute the functional requirements
described in the previous section. They are listed as non-functional
requirements as follows:

\begin{itemize}
  \item The DSP Data Persistence component must maintain the received DSP
  Measurement Messages in-memory for a specific period of time in order to
  decrease the write load on the database;
  \item The Persistence storage system must be able to cope with dozens of DSP
  Messages received from different sensors;
  \item The Persistence storage system must be able to save measurements quickly;
  \item The Persistence storage system must be able to scale without too much
  architectural changes, or schema changes;
  \item The Persistence storage system must not impose the use of
  database languages such as SQL \cite{sql} or XML XPath \cite{xml-xpath}
  related to the data being persisted. Instead, data must be handled through API
  calls, as better abstractions for non-computer science researchers
  \cite{sn-programming-language}.
\end{itemize}

@@

As described in the previous chapter, the data model chosen to support
persistence of collected data from sensor networks is the Key-Value-Pair Data
Model \cite{db-kvp}. Beyond the data collected from the sensor devices, Data
Provenance techniques must also be considered to better describe the collected
data \cite{sn-provenance}. In general, the data model must include the
following metadata besed on Data Provenance:

\begin{itemize}
  \item \textbf{Locality}: properties that describe the location of the data.
  For example, the host machine name or IP address, the GPS coordinate
  \cite{gps};
  \item \textbf{Identity}: what uniquely identities the data among the others
  produced by other sensor devices in the network, as for instance a
  UUID\footnote{Universally Unique Identifier};
  \item \textbf{Time-Dimensions}: when the data was ``sensed''
  \cite{db-provenance} and when the collection transaction happened
  \cite{sn-time-series}.
\end{itemize}

After analyzing the DSP Platform and the structure of the DSP Messages in
chapter \ref{netbeams-architecture}, the following set of keys are expected to
be part of the design of the data model. Any key represented by time or
location are related to the provenance support discussed in section
\ref{sec:sn-data-description}:

\begin{itemize}
  \item \textbf{sensor.ip\underline{ }address}: it is extracted
  from the DSP Message Produce and identifies which sensor generated the sampling;
  \item \textbf{sensor.location.latitude, longitude}: the latitude and
  longitude GPS coordinates of the position of the sensor device, captured from
  references of the device. This is currently provided as a parameter in the
  bootstrap message for the component;
  \item \textbf{message\underline{ }id}: it's extracted from each of the
  messages contained in the DSP Message Container;
  \item \textbf{time.transaction}: it is extracted from the DSP
  message container creation time and it is used to identify when the
  transaction occurred;
  \item \textbf{time.fact}: it is extracted from the
  SondeDataContainer's date and time, and identifies the time in which the
  collected data occurred.
\end{itemize}

It is important to note that this set of metadata must occur in every
collected data from any sensor device. However, on the contrary of the
metadata, a different set of properties will occur to describe the sensed
properties of a given sensor, and for this reason, the following structure is
defined to accommodate such data:

\begin{itemize}
  \item \textbf{observation}: this key defines the values of the document, and
  will have every different property of the sensor. For instance, the data key
  for the properties of the YSI device will contain the set of properties 
  defined for the device, such as battery, temperature, salinity, etc.
\end{itemize}

@@

Considering the business analysis described in section
\ref{sec:business-process-analysis}, the addition of a data persistence layer
to the DSP Platform can be achieved by adding Data Consumer (DC) as described
in section \ref{netbeams-architecture}. Based on the Requirements
specification described earlier, this component must be responsible for
providing the processing of the DSP Measure Messages, extracting the payload of
the body of the message and converting it to the data model defined in the
previous section. Finally, the data can be transmitted to the chosen database
system. Figure \ref{fig:NetBEAMS-Persistence-Server-Node-Components} provides
an architectural view of the DSP Server node with the loosely-coupled
components, highlighting the inclusion of the DSP Data Persistence.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/NetBEAMS-Persistence-Server-Node-Components}
  \caption{UML Components diagram of the DSP Data Collector}
  \label{fig:NetBEAMS-Persistence-Server-Node-Components}
\end{figure}

As it is depicted in figure
\ref{fig:NetBEAMS-Persistence-Server-Node-Components}, the DSP Data
Persistence component can be added into the system as an independent
plug-and-play component. In addition, the database system and the necessary
drivers are also added without any changes to the existing DSP infrastructure.
First, as DSP Messages are delivered to the server through the DSP Wire
Transport Server, as shown by flow 1, the DSP Platform is responsible to
forward a copy of the Measurement messages directly to the DSP Data
Persistence as shown by flow 2. Then, the sole responsibility of the DSP Data
Persistence is to transfer the received measurement messages to the database
system by using the connection drivers, as depicted by flow 3.

@@

The new proposed Data Persistence component must follow the DSP requirements
for a new DSP Component, which specifies the design-pattern of Data
Producer/Consumer. In a nutshell, the design of a DSP Component maintains the
dependency among 3 different major Java packages: it depends on 2 existing
packages, namely the OSGi Framework and the DSP Platform, as well as and an
additional package for the database access driver API, as shown in the UML
Package Dependency Diagram \cite{uml} of Figure
\ref{fig:DSP-Data-Persistence-Packages-Dependency}.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-Data-Persistence-Packages-Dependency}
  \caption{UML Components diagram of the DSP Data Collector}
  \label{fig:DSP-Data-Persistence-Packages-Dependency}
\end{figure}

The following sections details each step of the DSP Data Persistence activation
and parallel execution of the contained classes.

@@

The DSP Data Persistence component starts with an OSGi Activator class, and the
implementation of the DSP Component, resposible to provide services to the
platform. Based on these specifications, the following set of classes in
figure \ref{fig:DSP-DataPersistence-Activator-Class-Diagram} shows the UML
Class Diagram \cite{uml} implemented for the component.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-DataPersistence-Activator-Class-Diagram}
  \caption{UML Class diagram for the DSP Data Persistence Activator and Component}
  \label{fig:DSP-DataPersistence-Activator-Class-Diagram}
\end{figure}

The DSP Data Persistence Activator is the main class responsible for the
activation of the DSP Data Persistence component, as it extends the OSGi class
BundleActivator and maintains a reference of the DSP Data Persistence class.
Furthermore, the activator maintains references to the classes BundleContext
and ServiceReference both from the OSGi Framework. The former is responsible
for maintaining the link with the OSGi Framework, while the latter is used to
register the DSP Component as a service. Therefore, it is clear that the class
DSP Data Persistence Activator is the main communication interface between the
DSP Platform and the OSGi Framework.

While the DSP Data Persistence Activator is responsible for the orchestration
of the classes during the execution on the OSGi Framework, the class DSP Data
Persistence extends from the DSP Platform class DSPComponent. For this reason,
this class inherits all the behavior described on \ref{netbeams-architecture}
and can be in the roles of Data Consumer and a Data Producer.

\begin{itemize}
  \item \textbf{DSP Data Persistence as Data Consumer}: receives any measurement
  message from remote sensors;
  \item \textbf{DSP Data Persistence as Data Producer}: transforms any
  measurement message received into a format that must be ready to be saved on the
  database, based on the description of the data model in section
  \ref{sec:dsp-persistence-data-model}.
\end{itemize}

The behavior of the activator class follows the specifications of the
OSGi Framework and, for this reason, instances of the DSP Data Persistence
Activator can be in the states described on the UML State Diagram depicted in
Figure \ref{fig:DSPPlatform-Install-Usage-State-Diagram}. It is important to
note that the activator is an extension of the OSGI class BundleActivator, as
well as being responsible for stopping or starting the the DSP Data Persistence
component instance. In this way, following the specifications of section
\ref{netbeams-architecture}, the DSP Data Persistence Platform can be
initialized as described as it is described by the UML Sequence Diagram of
Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}:

\begin{enumerate}
  \item During the DSP Platform activation, it will get the name of the
  persistence component from the configuration artifact config.xml as seen in
  listing \ref{file:dsp-config.xml};
  \item After being selected based on the configuration priority, the DSP
  Bundle artifact is identified and installed, by creating an instance of the
  class DSP DataPersistence Activator and making a call to the method start();
  \item During the activation, an instance of the DSP Data Persistence
  component is created and registered as an OSGi Service;
  \item Upon registering the DSP Data Persistence component, the DSP Platform
  "listens" the event serviceUpdate() and, and triggers the last operations to
  bootstrap the DSP Component:
   \begin{enumerate}
      \item Initializes the component by calling the method initComponent();
      \item Starts the component by calling the method startComponent();
      \item Bootstrap messages are delivered to the component.
   \end{enumerate}
\end{enumerate}

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as shown
in Listing \ref{file:dsp-data-persistence-bootstrap.xml}. The initialization of
the DSP Component Activator and the DSP Component registration are also shown
in figure \ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}
  \caption{UML Sequence diagram describing the DSP Data Persistence bundle activation}
  \label{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}
\end{figure}

As soon as the gate "Start DSP Activator Gate" reaches the DSP Data
Persistence instance, it first creates the unique instance of the class DSP
Data Persisence and uses that instance to register it as the OSGi service
through the call to the DSP Platform helper class ActivatorHelper. At this
point, the DSP Data Persistence has been installed and is in the state
"Active".

@@

Although the DSP Data Persistence component has been instantiated as of image
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}, 
the process of delivering messages to the DSP Data Persistence has yet to be
completed. As defined by the requirements, the DSP Component must execute its
function in a concurrent fashion, using configuration parameters during its
bootstrap process. The class DSP Data Flusher is responsible for running as a
"worker thread", whose function is to verify if there are messages to be
flushed into the Database with the class Transient Persistence Layer. Figure 
\ref{fig:DSP-DataPersistence-Flusher-Classes} describes the UML Class
Diagram \cite{uml} for the main classes of this functionality.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-DataPersistence-Flusher-Classes}
  \caption{UML Class diagram with the DSP Data Flusher worker thread}
  \label{fig:DSP-DataPersistence-Flusher-Classes}
\end{figure}

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/DSP-Data-Persistence-Mongo-Classes}
  \caption{UML Class diagram with Persistence Service for MongoDB}
  \label{fig:DSP-Data-Persistence-Mongo-Classes}
\end{figure}

>>>>

The mongoDB database system was selected, as stated in chapter 5, to store 
the collected data from the SF-BEAMS sensor network. NetBEAMS is based on a 
component-based architecture, where each of the components can play different
roles during their lifecycle. It uses the abstraction of Data Producers and 
Data Consumers: while a component produces data to the system, the latter 
consumes \cite{netbeams2009}. In this way, a DSP Data Component was designed
and implemented to provide the integration between NetBEAM and the mongoDB 
database. This integration is based on the Data Sensor Platform (DSP), the
kernel software stack provided by NetBEAMS, as documented in
\cite{netbeams-dsp-architecture}. Finally, details about the implementation of
this design are located in the Appendix Section
\ref{sec:dsp-data-persistence-implementation}.

@@

As described in \cite{netbeams-dsp-architecture}, the DSP Platform was designed
to support SF-BEAMS. As shown in Figure \ref{fig:dsp-persistence-business-process}, 
the event of data collection starts by the DSP Component deployed on each 
DSP Gateway Node \cite{netbeams-dsp-architecture}, by the event ``Measurements
X Collected''. When this DSP Client component finishes the data extraction
process, a new one starts. First, each of the co-located DSP Client processes 
the collected measurements and creates a DSP Message, embodying the collected 
data from the associated sensor device. Each DSP Message is added into a local
temporary outbound queue that waits to be transmitted. Then, an asynchronous 
data transmission event occurs at a specified rate, packaging all DSP
Messages into the body of a DSP Messages Container in order to be serialized
into XML (shown in Listing \ref{stream:dsp-message-serialized-ysi}) and
transferred to the DSP Server host, a centralized server that receives all the
collected data embodied into each of the DSP Messages.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.6]{../diagrams/DSP-DataPersistence-Business-Diagram}
  \caption{UML Business Process Model diagram - Adding Persistence for NetBEAMS}
  \label{fig:dsp-persistence-business-process}
\end{figure}

The goal of the new DSP Data Persistence component is to proxy the payload 
from the DSP Messages to the new suggested Measurements Database. Once the DSP
Server receives all the messages in a container, it processes the messages and
sends their payloads to the Measurements Database. This process requires a new
Database Configuration to be provided to the DSP Server host, as well as the
installation and setup of the new Measurements Database. In this way, different
users of the collected data, such as Programmers and Biologists, can access the
collected data directly from the Measurements Database and reuse them by
Exporting specific observation parameters to different formats of their
interest.

@@

This section outlines the functionalities and constraints that the developed
solution must perform and follow, respectively, obtained by analyzing the main
data collection scenario described in the previous section and reviewing the
requirements defined by the RTC staff members.

@@

The primary goal of the system is to allow users to have easy access to the 
collected data from \emph{NetBEAMS} using a schema-less database technology.
mongoDB provides access to the stored data without requiring the use of 
a Structured Query Language. Instead, that technology provides a programming
language abstraction that makes the access to the stored data easier 
\cite{sn-programming-language}. Using Agile User Stories to describe Use 
Cases Diagrams \cite{use-cases-user-stories}, Figure 
\ref{fig:DSP-Data-Persistence-UseCases-Diagram-Users} enumerates these
functionalities as customers conversations, starting with a fictitious
persona, the action it desires to perform and the outcome of the action, a
requirements gathering pattern defined in Scrum \cite{agile-scrum}:

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.65]{../diagrams/DSP-Data-Persistence-UseCases-Diagram-Users}
  \caption{UML Use Case diagram for Persistence Functions}
  \label{fig:DSP-Data-Persistence-UseCases-Diagram-Users}
\end{figure}

\begin{itemize}
  \item \textbf{Search Measurements by Sensor Properties}: "As a marine
  biologist, I would like to search observations by filtering values of the 
  sensor device's properties such as temperature or salinity, so that I can 
  find specific associated values to the observation.";
  \item \textbf{Search Measurements by Sensor Location}: "As an oceanographer,
  I would like to search observations that took place at the geographic
  coordinates (x latitude, y longitude), so that I can assess the area around
  the given coordinates.";
  \item \textbf{Search Measurements by Data collection date}: "As a marine 
  biologist, I would like to search observations that took place last week, 
  so that I can assess past environmental conditions";
  \item \textbf{Annotate Existing Measurements}: "As a estuarine ecologist,
  I would like to annotate observations from the time the oil spill occurred
  in the San Francisco Bay, so that I can maintain historical evidence of 
  the impact of such event.";
  \item \textbf{Export Data to Spreadsheets}: "As a scientist from RTC, I 
  would like to analyze of the observed data from yesterday using a 
  spreadsheet, so that I can verify measurements using Microsoft Excel.";
  \item \textbf{Export Data to OPeNDAP format}: "As a biologist, I would 
  like to export the collected data produced during this month using the
  OPeNDAP data format, so that I can collaborate with other research groups
  that uses this data format.";
  \item \textbf{Access the data through API, Programming Languages}: "As a
  marine biologist who learned the Python scripting language, I would like to
  write a software that connects directly to the database, so that I can
  implement a customized analysis system for NetBEAMS.";
  \item \textbf{Remove Undesired Measurements}: "As an oceanographer, I
  would like to remove specific observations collected yesterday, 
  so that the research group does not use 'junk data'.".
\end{itemize}

As one can see, the basic users' functionalities are related to the access to
the collected data using different parameters related to the each observation's
properties or implied metadata such as time of collection or location. In order
to do so, the functions to be performed were developed following the
specifications of the Data Sensor Platform (DSP) architecture, as described in
\cite{netbeams-dsp-architecture}. In order to collect data sent to the NetBEAMS
server, the new DSP component must act as a Data Consumer. The functionalities
provided by this component can be seen in the UML Use Cases diagram \cite{uml}
in Figure \ref{fig:DSP-Data-Persistence-UseCases-Diagram-System}, and
summarized as follows:

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/DSP-Data-Persistence-UseCases-Diagram-System}
  \caption{UML Use Case diagram for the Persistence Component}
  \label{fig:DSP-Data-Persistence-UseCases-Diagram-System}
\end{figure}

\begin{itemize}
  \item \textbf{Update Component Properties}: the component must be
  configurable, with its own set of properties, such as the flush rate used 
  to send data to the database system;
  \item \textbf{Update Database Properties}: the component must be able to
  update the mongoDB database settings such as IP Address and  communication 
  port;
  \item \textbf{Collect DSP Measurement Message}: the component must be
  able to receive and temporarily retain the collected messages from
  different the sensors devices;
  \item \textbf{Categorize Measurement Message}: the component must be able
  to categorize the DSP Message, as it arrives, by its properties such as 
  transaction time, and originating DSP Message ID;
  \item \textbf{Save Messages into the Database}: the component must be able
  to send the payload of the DSP Messages to the database.
\end{itemize}

@@

Different constraints were required to execute the functional requirements
described in the previous section. Those constraints are listed as different
non-functional requirements as follows:

\begin{itemize}
  \item The solution must maintain the received DSP Measurement Messages
  in-memory for a specific period of time in order to decrease the write load
  on the database (flush rate);
  \item The persistence storage system must be able to cope with the Data
  Volume produced by SF-BEAMS on daily basis;
  \item The persistence storage system must be able to scale without minimal
  architectural and schema changes;
  \item The persistence storage system must be able to provide access to the
  persisted data using an programmatic abstraction interface, as the authors
  of \cite{sn-programming-language} proposed.
\end{itemize}

Driven by the requirements defined, the foundation of the integration between
the DSP Data Persistence component and the mongoDB database system is the
specification of the Keys that comprise a set of observations collected from
the sensor devices.

@@

MongoDB is defined as a Document-Oriented Database system, which is based on 
the abstraction of the Key-Value-Pair data model, using a binary version of JSON
\cite{json} as its main data representation. MongoDB does not require the creation 
of a database schema, rather the definition of key descriptions that comprises
a document on the database organized into separate collections of relating
documents. In order to describe the keys of a document, the specifications
detail the use of Data Provenance as guidelines to the description of the
collected data. Therefore, the key-value pairs of any instance of the
collected data must include the following key properties:

\begin{itemize}
  \item \textbf{Location}: properties that describe the location of the data.
  For example, the host machine name, IP address, or the GPS coordinates
  \cite{gps} from the sensor device;
  \item \textbf{Identity}: what uniquely identifies the data among 
  others, produced by other sensor devices in the network such as the DSP
  Message ID using UUID\footnote{Universally Unique Identifier};
  \item \textbf{Time-Dimensions}: when the data was recorded 
  \cite{db-provenance} and when the transaction occurred \cite{sn-time-series}.
\end{itemize}

After analyzing the DSP Platform and the structure of the DSP Messages in
\cite{netbeams-dsp-architecture}, the values of the collected data must be
categorized into two different sets of data: the ``Key Segment'' and the
``Value Segment''. The key segment provides the necessary metadata 
used to describe the collected data. An instance of the design of the 
keys is shown in Listing \ref{file:mongodb-ysi-data-format}:

\begin{itemize}
  \item \textbf{message\underline{ }id}: Defines the DSP Message ID that
  carried the observation from the sensor device to the DSP Server. This ID
  can be used during network and database inspections in order to track messages
  for the DSP Platform;
  \item \textbf{sensor}: Defines the properties of the sensor device;
    \begin{itemize}[label=\textbullet]
        \item \textbf{ip\underline{ }address} Defines the IP Address of the 
        sensor device that produced the data;
        \item \textbf{location}: Defines the geographical coordinates of the
        sensor device, using the Decimals Degree format \cite{decimal-degrees};
            \begin{itemize}[label=\textbullet]
                \item \textbf{latitude} the GPS latitude coordinate of the
                position of the sensor device;
                \item \textbf{longitude}: the GPS longitude coordinate of the
                position of the sensor device.
            \end{itemize}
    \end{itemize}
  \item \textbf{time}: Defines different time dimensions according to Data 
   Provenance;
    \begin{itemize}[label=\textbullet]
      \item \textbf{valid}: it is extracted from the container's date and time
      of creation, which was collected during the sensor device's reading;
      \item \textbf{transaction}: it is extracted from the DSP message
         container's creation time and it is used to identify when the
         transaction, which transferred the message from the DSP Gateway Client
         to the DSP Server, occurred;
    \end{itemize} 
\end{itemize}

An associated ``Value Segment'' for the aforementioned ``Key Segment'' carries
the observations of a given sensor device. The key \textbf{observation} stores
the values of the collected data. For instance, the observation key of
for the YSI device \cite{YSI-Sonde} may contains all the subsets of 
the properties defined for the device, such as battery, temperature and
salinity, among others. An example of such use of the segments is shown in
Listing \ref{file:mongodb-ysi-data-format}.

Based on this foundation, the design of the DSP Data Persistence component
evolved from a top-down approach.

@@

Considering the business analysis described in Section
\ref{sec:business-process-analysis}, the addition of a data persistence layer
to the DSP Platform can be achieved by adding a Data Consumer (DC) as 
described by \cite{netbeams-dsp-architecture}. Based on the 
Requirements specification, this component must be responsible for the 
extraction process that takes the payload of the body of a filtered DSP
Measurement Message and converts it to the key-value data model defined
in the previous section. This process guarantees that the collected data is
inserted to a running mongoDB database system and gives covers the use cases
defined. Figure \ref{fig:NetBEAMS-Persistence-Server-Node-Components}
provides an architectural overview of the DSP Server node with the
loosely-coupled components, highlighting the inclusion of a new DSP Component
called ``DSP Data Persistence''.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/NetBEAMS-Persistence-Server-Node-Components}
  \caption{UML Components diagram of the DSP Data Collector}
  \label{fig:NetBEAMS-Persistence-Server-Node-Components}
\end{figure}

As it is depicted the diagram 
\ref{fig:NetBEAMS-Persistence-Server-Node-Components}, the DSP Data
Persistence can be added into the existing architecture as an 
independent plug-and-play component. The database system and the necessary
drivers are also added without any changes to the existing DSP 
infrastructure. First, the DSP Messages are delivered to the DSP server 
through the DSP Wire Transport Server, as shown by flow 1. Next, the DSP 
Platform is responsible to forward a copy of the Measurement messages 
directly to the DSP Data Persistence as shown by flow 2. Then, the sole 
responsibility of the DSP Data Persistence is to transfer the received 
measurement messages to the database system by using the connection 
drivers, as shown by flow 3.

Any DSP Component is essentially an OSGi Bundle \cite{osgi-intro}, a unit of
deployment in a Java JAR format \cite{java-tutorial} containing the Java
classes and a descriptor manifest artifact that identifies the major components
of the package, which services the component requires to use and which ones it
will optionally publish during its activation and lifecycle. Details about the
deployment of the DSP Data Persistence component, as well as about the OSGi
environment used, are given in the Appendix section
\ref{sec:dsp-component-osgi-deployment}.

@@

The proposed Data Persistence component must follow the DSP Component
requirements, which specifies the design-pattern of Data Producer and Consumer
mentioned earlier in this chapter. As seen in the UML Package  Dependency
Diagram in Figure \ref{fig:DSP-Data-Persistence-Packages-Dependency}, a new
DSP Component has dependency on three different packages: the OSGi Framework
and the DSP Platform responsible to integrate a new component into the
micro-kernel, and the mongoDB Java driver API that connects to the mongoDB
server.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/DSP-Data-Persistence-Packages-Dependency}
  \caption{UML Components diagram of the DSP Data Collector}
  \label{fig:DSP-Data-Persistence-Packages-Dependency}
\end{figure}

In order to use a DSP Component, one must first activate it into the OSGi
Framework the DSP Platform runs on. Both the activation and the design of how
the new component acts as a Data Consumer is explained in the following
sections.

@@

The DSP Data Persistence component starts with an OSGi Activator class, and 
the implementation of the DSP Component, responsible for providing services 
to the platform \cite{netbeams-dsp-architecture}. Based on these 
specifications, the following set of classes, shown in Figure 
\ref{fig:DSP-DataPersistence-Activator-Class-Diagram}, depicts the implemented
classes for the new DSP Data Persistence component.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/DSP-DataPersistence-Activator-Class-Diagram}
  \caption{UML Class diagram for the DSP Data Persistence Activator and Component}
  \label{fig:DSP-DataPersistence-Activator-Class-Diagram}
\end{figure}

The class DSPDataPersistenceActivator is the main class responsible for the
activation of the class DSPDataPersistence, as it extends from the OSGi
class BundleActivator. The BundleActivator maintains references to the classes
BundleContext and ServiceReference, both from the OSGi Framework, in order to
maintain the component managed by the OSGi Framework and register the class
DSPDataPersistece as a service to the OSGi environment. Therefore, the class
DSPDataPersistenceActivator is the main communication interface between the 
DSP Platform and the OSGi Framework, are responsible for the orchestration of
the application lifecycle.

The class DSPDataPersistence implements the interface DSPComponent, responsible
to provide the contract between the component and the DSP platform. As a
consequence, this class inherits all the behavior described in
\cite{netbeams-dsp-architecture} necessary to be executed in the DSP Platform
in a given context (through the instance of class DSPContext). The different
roles played by the class DSPDataPersistence are defined as follows:

\begin{itemize}
  \item \textbf{Data Consumer}: receives any measurement message from remote
  sensors;
  \item \textbf{Data Producer}: transforms any received measurement message
  into the format defined in Section \ref{sec:dsp-persistence-data-model}, in
  order to be sent to the database.
\end{itemize}

The class DSPDataPersistenceActivator is responsible for starting and stopping,
the services provided by the DSPDataPersistence component, as this behavior is
inherited from the OSGi class BundleActivator from OSGi Framework. Following
the specification of the DSP Architecture \cite{netbeams-dsp-architecture},
Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram} 
shows a UML Sequence Diagram that depicts the steps of the activation of the
DSP component, and summarized as follows:

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}
  \caption{UML Sequence diagram describing the DSP Data Persistence bundle activation}
  \label{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}
\end{figure}

\begin{enumerate}
  \item During its activation, the DSP Platform will get the name of the
  DSP Data Persistence component from the configuration artifact config.xml as
  seen in Listing \ref{file:dsp-config.xml};
  \item After being selected based on the configuration priority, the 
  DSPBundle artifact is identified and installed by creating an instance of the
  class DSPDataPersistenceActivator, making a call to the method start();
  \item During the activation, an instance of the class DSPDataPersistence
  is created and registered as an OSGi Service;
  \item Upon registering the DSP Data Persistence component, the DSP Platform
  listens for the event serviceUpdate() and triggers the last operations to
  bootstrap the DSP Component:
   \begin{enumerate}
      \item Initialize the component by calling the method initComponent();
      \item Start the component by calling the method startComponent();
      \item Bootstrap messages by calling the method delivery(). This is where
      both the DSP Measurements Messages and Update Messages are delivered by
      the DSP Broker. This concept is based on the listener of Observer
      Design-Pattern \cite{gof}.
   \end{enumerate}
\end{enumerate}

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as shown
in Listing \ref{file:dsp-data-persistence-bootstrap.xml}. The initialization of
the DSP Component Activator and the DSP Component registration are also shown
in Figure \ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

As soon as the process "Start DSP Activator Gate" is executed, an instance of
the class DSPDataPersistence component is created to be registered into the
OSGi Framework using the DSP Platform's helper class ActivationHelper. It is
responsible to proxy the method call to the class BundleContext, which makes
the actual registration of the DSPDataPersistence component as an OSGi
Service. At this point, the DSPDataPersistence is installed and on the state
"Active", as defined by the OSGi Framework (Refer to
\cite{netbeams-dsp-architecture} for details).

Once the DSP Data Persistence component is concurrently running with other
components from both the DSP Platform and the OSGi Framework, it is able to
receive DSP Messages. The following section will explain how to bootstrap the
DSP Data Component in order to receive Measurement Messages.

@@

Although the DSP Data Persistence component has been instantiated as of Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}, 
the process of delivering messages to the DSP Data Persistence component has yet
to be completed. As defined by the requirements, the DSP Component must execute its
function concurrently, using the configuration parameters during its
bootstrap process. The DSP Platform delivers any bootstrap message defined in
the DSP Platform runtime environment, as explained in the Appendix section
\ref{sec:dsp-persistence-boostrap}. However, to deliver DSP Measurement
Messages, a new filter must be provided. As explained in
\cite{netbeams-dsp-architecture}, the DSP Broker uses the DSP Matcher's rules
to select to Data Consumers configured in a given DSP. For this reason, a rule
that filters any remote measurement message received by running DSP Server is
used, as shown in Listing \ref{file:dsp-matcher-config.xml}. The rule defined
in line 63 express that all messages created by any producer on any DSP Host,
either client or sever, has as delivery target the DSP Data Persistence
component. In this way, when DSP Measurement Messages are delivered, the DSP
Broker delivers the DSP Message to the DSP Data Persistence Component.

The UML Class Diagram for the participating classes of the DSP Data
Persistence component is shown in Figure
\ref{fig:DSP-DataPersistence-Flusher-Classes}. The class DSPDataFlusher is
responsible for running as a worker thread, whose function is to verify if
there are messages to be flushed to the Database by contacting the Singleton
class \cite{gof} TransientPersistenceLayer, responsible to keep track of the
DSP Messages received when the DSP Broker delivered them to the DSP Data
Persistence component.
